The effects of Water and Sanitation on Tuberculosis Incidence in South East Asian and African countries

### 2025-04-21 products/manuscript/Manuscript.qmd
Fixing render issues

---
title: "The effects of Water and Sanitation on Tuberculosis Incidence in South East Asian and African countris"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: assets/dataanalysis-references.bib
csl: assets/american-journal-of-epidemiology.csl
editor: 
  markdown: 
    wrap: 72
---

This uses MS Word as output format. [See
here](https://quarto.org/docs/output-formats/ms-word.html) for more
information. You can switch to other formats, like html or pdf. See [the
Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
install.packages("dplyr")
library(dplyr)
install.packages("gt")
library(gt)
```

@@

-   Muhammad Nasir$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
-   Second Author$^{3, *}$
-   Third Author$^{3, \dagger}$
-   Last Author$^{1, 2, \land}$

@@

# Summary/Abstract

*Write a summary of your project.*

@@

#| label: fig-1
#| fig-cap: "Trends of Tuberculosis of each country from 2000 to 2020"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "eda", "tbi_spageti.png"))
```

@@

```{r}
#| label: fig-2
#| fig-cap: "Trends of All Variables over the time 2000 to 2020 by countries"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "eda", "plot_all_vars.png"))
```

```{r}
#| label: fig-3
#| fig-cap: "Trends of All Variables over the time 2000 to 2020 by countries"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "eda", "plot_all_vars_region.png"))
```

@@

#| label: fig-4
#| fig-cap: "Non-linear relationhsip between Tuberculosis Incidence and Predictors "
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "plots_tbi_vs_allvars.png"))

@@

```{r}
#| label: fig-5
#| fig-cap: "Non-linear relationhsip between Tuberculosis Incidence and Predictors after transfirmation "
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "plot_transformed.png"))
```

@@

#| label: fig-6
#| fig-cap: "Spearman Correlation Matrix "
#| echo: FALSE
knitr::include_graphics(here("results","figures", "eda", "correlation_matrix.png"))

@@

#| label: fig-7
#| fig-cap: "AIC and RMSE "
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "plot_aic_rmse.png"))
```

@@

We fitted a generalized additive mixed model (GAMM) using a negative binomial distribution with a log link to model tuberculosis incidence (TBI) rate across countries in South East Asia and Africa. The model incorporated smooth functions of year, access to basic and safe water and sanitation services, and the Human Development Index (HDI), as well as interaction terms between key environmental and socioeconomic indicators. A random effect was included for country to account for unobserved heterogeneity. (@tbl-1 ) shows the result of the final model. 
```{r}
#| label: tbl-1
#| tbl-cap: "Summary of GAMM Parametric and Smooth Terms"
#| echo: false
#| message: false
#| warning: false

@@

gamm_tbl %>%
  gt(groupname_col = "type") %>%
  fmt_number(columns = where(is.numeric), decimals = 3) %>%
  tab_header(
    title = "Summary of GAMM Estimates",
    subtitle = "Parametric and Smooth Terms"
  ) %>%
  cols_label(
    term = "Term", estimate = "Estimate", std.error = "Std. Error",
    edf = "EDF", chi_sq = "Chi-square", p.value = "p-value"
  ) %>%
  fmt_missing(everything(), missing_text = "-")

@@

```{r}
#| label: fig-8
#| fig-cap: "Predicted vs Observed TBI originalvs generated data "
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "plot_ob_pre_ml_vs_ori.png"))
```

@@

```{r}
#| label: fig-9
#| fig-cap: "Predicted and observed value for Xgboost using train and test data"
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "plot_obs_pred_ml.png"))
```

@@

```{r}
#| label: fig-10
#| fig-cap: "Tunning Result for XGBoost"
#| echo: FALSE
knitr::include_graphics(here("results","figures", "analysis", "autoplot_xgb_tuning.png"))
```

@@

Our findings have several policy implications. First, they highlight the need for multisectoral strategies that address both environmental conditions and broader socioeconomic development to combat TB effectively. Investments in water infrastructure should be accompanied by programs targeting education, nutrition, housing, and healthcare access [@onozaki2015impact; @udwadia2014multidrug]. Second, the complex interactions observed suggest that context-specific interventions are essential. A one-size-fits-all approach may fail to capture the unique synergy between environmental and developmental factors within each country.

@@

Despite these limitations, this study adds to the growing evidence that TB incidence is deeply embedded in broader developmental and environmental systems. Our findings support the WHO End TB Strategy, which emphasizes tackling the social determinants of TB alongside biomedical interventions [@who2015end; @hargreaves2011social].

>>>>

---
title: "The effects of Water and Sanitation on Tuberculosis Incidence in South East Asian and African countries"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
execute:
  echo: false
bibliography: assets/citations.bib
csl: assets/american-journal-of-epidemiology.csl
editor:
  markdown:
    wrap: 72
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
#install.packages("dplyr")
library(dplyr)
#install.packages("gt")
library(gt)
```

@@

-   Muhammad Nasir$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)

@@

# Abstract

Tuberculosis (TB) remains a significant global health concern, particularly in low- and middle-income countries. In 2023, an estimated 10.8 million people contracted TB, with nearly 70% of cases reported in South-East Asia and Africa. While TB is primarily airborne, environmental and socioeconomic factors, including water and sanitation access, may influence transmission by affecting hygiene and immune function. This study aims to investigate the impact of access to water and sanitation services on TB incidence across South-East Asia and Africa, incorporating the Human Development Index (HDI) as a control for socioeconomic context. A Generalized Additive Mixed Model (GAMM) was applied to longitudinal data from 2000 to 2022. The model included smooth terms for year, water and sanitation indicators, and HDI, with country as a random effect.The model showed excellent fit (adjusted R² = 0.997), with significant non-linear effects observed for year, access to basic water (p = 0.013), safe water (p < 0.001), and HDI (p < 0.001). Sanitation variables were not independently significant. However, key interactions—such as between water access and HDI—were highly significant (p < 0.001), suggesting complex environmental-development dynamics influencing TB risk.In conclusion, expanding water access and addressing development disparities may be critical in reducing TB burden in high-incidence regions.

@@

#| label: fig-1
#| fig-cap: "Trends of Tuberculosis of each country from 2000 to 2020"
#| echo: false
knitr::include_graphics(here("results", "figures", "eda", "tbi_spageti.png"))

@@

#| label: fig-2
#| fig-cap: "Trends of All Variables over the time 2000 to 2020 by countries"
#| echo: false
knitr::include_graphics(here("results", "figures", "eda", "plot_all_vars.png"))

#| label: fig-3
#| fig-cap: "Trends of All Variables over the time 2000 to 2020 by countries"
#| echo: false
knitr::include_graphics(here("results", "figures", "eda", "plot_all_vars_region.png"))

@@

#| label: fig-4
#| fig-cap: "Non-linear relationhsip between Tuberculosis Incidence and Predictors "
#| echo: false
knitr::include_graphics(here("results","figures", "analysis", "plots_tbi_vs_allvars.png"))

@@

#| label: fig-5
#| fig-cap: "Non-linear relationhsip between Tuberculosis Incidence and Predictors after transfirmation "
#| echo: false
knitr::include_graphics(here("results","figures", "analysis", "plot_transformed.png"))

@@

#| label: fig-6
#| fig-cap: "Spearman Correlation Matrix "
#| echo: false
knitr::include_graphics(here("results","figures", "eda", "correlation_matrix.png"))

@@

#| label: fig-7
#| fig-cap: "AIC and RMSE "
#| echo: false
knitr::include_graphics(here("results","figures", "analysis", "plot_aic_rmse.png"))

@@

We fitted a generalized additive mixed model (GAMM) using a negative binomial distribution with a log link to model tuberculosis incidence (TBI) rate across countries in South East Asia and Africa. The model incorporated smooth functions of year, access to basic and safe water and sanitation services, and the Human Development Index (HDI), as well as interaction terms between key environmental and socioeconomic indicators. A random effect was included for country to account for unobserved heterogeneity. (@tbl-1 ) shows the result of the final model. 
```{r}

#| label: tbl-1
#| tbl-cap: "Summary of GAMM Parametric and Smooth Terms"
#| echo: false
#| message: false
#| warning: false

@@

gamm_tbl %>%
  gt(groupname_col = "type") %>%
  fmt_number(columns = where(is.numeric), decimals = 3) %>%
  tab_header(
    title = "Summary of GAMM Estimates",
    subtitle = "Parametric and Smooth Terms"
  ) %>%
  cols_label(
    term = "Term", estimate = "Estimate", std.error = "Std. Error",
    edf = "EDF", chi_sq = "Chi-square", p.value = "p-value"
  ) %>%
  sub_missing(everything(), missing_text = "-")

@@

```{r}
#| label: fig-8
#| fig-cap: "Predicted vs Observed TBI originalvs generated data "
#| echo: false
knitr::include_graphics(here("results","figures", "analysis", "plot_ob_pre_ml_vs_ori.png"))
```

@@

#| label: fig-9
#| fig-cap: "Predicted and observed value for Xgboost using train and test data"
#| echo: false
knitr::include_graphics(here("results","figures", "analysis", "plot_obs_pred_ml.png"))

@@

#| label: fig-10
#| fig-cap: "Tunning Result for XGBoost"
#| echo: false
knitr::include_graphics(here("results","figures", "analysis", "autoplot_xgb_tuning.png"))

@@

Our findings have several policy implications. First, they highlight the need for multisectoral strategies that address both environmental conditions and broader socioeconomic development to combat TB effectively. Investments in water infrastructure should be accompanied by programs targeting education, nutrition, housing, and healthcare access [@onozaki2015impact]. Second, the complex interactions observed suggest that context-specific interventions are essential. A one-size-fits-all approach may fail to capture the unique synergy between environmental and developmental factors within each country.

@@
